Linear Stochastic Approximation- Constant Step-Size and Iterate Averaging

### 2017-05-13 stepsizes.tex
refreshing

\section{Automatic Tuning of Step-Sizes}
\begin{comment}
We now argue, in a heuristic manner, as to why constant step-size with PR-averaging might be a principled idea, especially in TD type algorithms. Consider the following LSA with time varying step-sizes $\alpha_t>0$
\begin{align}
\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1}).
\end{align}
It is known from asymptotic results by \cite{borkarbook,kushner}\todoch{Check about Kushner} that in order to handle the noise the step-sizes have to be diminishing i.e., $\alpha_t\ra$, unsummable i.e., $\sum_{t \geq 0}\alpha_t=\infty$ and square summable $\sum_{t\geq 0}\alpha_t^2<\infty$. We now look at two specific diminishing step-size sequences, to illustrate the issues that the diminishing step-sizes presents us with. In what follows, we assume that the maximum eigevalue of $\EE{A_t}$ is less that $1$.
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item \textbf{$\alpha_t=\frac{1}{t}$\footnote{While $\frac{c}{t}$, $c\geq 0$ is more general, it is enough to consider $\frac1t$ by accommodating the multiple $c$ as part of the matrix $A_t$}} is known yield a rate of $\frac{1}{t^{\lambda_{\min}}}$, where $\lambda_{\min}$ is the smallest real part of the eigen values of the matrix $\EE{A_t}$.
\item \textbf{$\alpha_t=\frac{c}{c+t}$} is another example, where the step size decays at a rate $\frac{1}{t/c}$. The rate of decay of error in this case turns out to be $\frac{c^2\sigma_b^2}{t}$, whenever $c>\frac{1}{\lambda_{\min}}$, which can be ensured only when $\lambda_{\min}$ is known.
\end{itemize}
In contrast, in the case of constant step-size with PR-averaging, not knowing the problem instance can result in instability, and when we manage to choose a step-size that results in stable behavior, the rates obtained in \Cref{th:pdrate,cor:as} hold for it. Using this observation, the constant-step size can be set in a more prinicipled manner as showing the pseudocode below.
\end{comment}
In \Cref{sec:uniform} we showed a negative result that for the general class of LSAs, it is not possible to pick a fixed uniform step-size that can be used for any instance of the class. Further,
recall from our discussion in \Cref{sec:mainresults} that as $\alpha\ra\alpha_P$ (or $\alpha\ra\alpha_{P_U}$), the MSE blows up and for $\alpha>\alpha_P$ the LSA is unstable because the spectral norm of the random matrix $I-\alpha A_t$ is no longer less that unity. Combining these facts, we conclude that if the step-size is incorrect for a particular instance then we would observe instability, a fact that can be used to reset the step-size to a smaller value, and this procedure can be continued in an iterative fashion to find the right choice of step-size for a given problem. We formalize this idea in the form of a step-size tuning algorithm in \Cref{alg:tuning}

\begin{algorithm}[H]
\caption*{Automatic Tuning of Constant Step-Size}
\begin{algorithmic}[1]
\STATE{Initialize: $\theta_0$, $\alpha=1$, $i=1$, $\rho_0=0$}
\FOR{$t=1,2,\ldots, $}
\STATE{$\theta_t=\theta_{t-1}+\alpha (b_t-A_t\theta_{t-1})$}
\IF{$t (mod k)==0$}
\STATE{$\rho_{i}=\norm{\theta_t}$}
\IF{IsUnstable($\{\rho_j,j=1,\ldots,i\}$)==1}
\STATE{$\alpha=\alpha/2$}
\ENDIF
\STATE{$i=i+1$}
\ENDIF
\ENDFOR
\end{algorithmic}
\label{alg:tuning}
\end{algorithm}
In \Cref{alg:tuning} IsUnstable$(\rho_j,j=1,\ldots,i\}$) is a routine whose input is $\norm{\theta_t}$ measured at time instances $k$ apart. It is easy to show from Gronwall type arguments that the $\norm{\theta_t}\leq (1-exp{\beta t})$, where $\beta>0$ implies stability and $\beta<0$ imples instability. From this Gronwall based inequality, we observe that when the iterates are stable the ratio $\frac{\rho_{k+1}}{\rho_k},k=1,\ldots,i-1$ should be `roughly' decreasing (because there is the iterates are persistently corrupted by noise) and should converge to $1$. This condition on the sequence $\frac{\rho_{k+1}}{\rho_k}$ can be checked in practice with appropriate tolerances.
\begin{comment}
In order to calculate the rates we make use of the following asymptotic result by \cite{borkarbook}.
\begin{theorem}\label{th:borkar}
Let $\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1})$, where $\theta_t,b_t\in\R^d$ and $A_t\in\R^{\dcd}$ that satisfy the conditions in \Cref{assmp:lsa} and $\sum_{t\geq 0}\alpha_t=\infty$, $\sum_{t\geq 0}\alpha^2_t<\infty$. Let $t_{ode}(k)=\eqdef \sum_{i=0}^{k-1}\alpha_n$ and let $\tilde{\theta}(t)$ be the trajectory of the ordinary differential equation $\dot{\theta}(t)=\EE{b_t}-A_P\theta(t)$. Then it follows that
\begin{align*}
\lim_{k\ra\infty}\sup\norm{\tilde{\theta_t(t_{ode}(k))}-\theta_t} \ra 0
\end{align*}
\end{theorem}

>>>>

\section{Automatic Tuning of Step-Sizes}
\begin{comment}
We now argue, in a heuristic manner, as to why constant step-size with PR-averaging might be a principled idea, especially in TD type algorithms. Consider the following LSA with time varying step-sizes $\alpha_t>0$
\begin{align}
\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1}).
\end{align}
It is known from asymptotic results by \cite{borkarbook,kushner}\todoch{Check about Kushner} that in order to handle the noise the step-sizes have to be diminishing i.e., $\alpha_t\ra$, unsummable i.e., $\sum_{t \geq 0}\alpha_t=\infty$ and square summable $\sum_{t\geq 0}\alpha_t^2<\infty$. We now look at two specific diminishing step-size sequences, to illustrate the issues that the diminishing step-sizes presents us with. In what follows, we assume that the maximum eigevalue of $\EE{A_t}$ is less that $1$.
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item \textbf{$\alpha_t=\frac{1}{t}$\footnote{While $\frac{c}{t}$, $c\geq 0$ is more general, it is enough to consider $\frac1t$ by accommodating the multiple $c$ as part of the matrix $A_t$}} is known yield a rate of $\frac{1}{t^{\lambda_{\min}}}$, where $\lambda_{\min}$ is the smallest real part of the eigen values of the matrix $\EE{A_t}$.
\item \textbf{$\alpha_t=\frac{c}{c+t}$} is another example, where the step size decays at a rate $\frac{1}{t/c}$. The rate of decay of error in this case turns out to be $\frac{c^2\sigma_b^2}{t}$, whenever $c>\frac{1}{\lambda_{\min}}$, which can be ensured only when $\lambda_{\min}$ is known.
\end{itemize}
In contrast, in the case of constant step-size with PR-averaging, not knowing the problem instance can result in instability, and when we manage to choose a step-size that results in stable behavior, the rates obtained in \Cref{th:pdrate,cor:as} hold for it. Using this observation, the constant-step size can be set in a more prinicipled manner as showing the pseudocode below.
\end{comment}
\begin{comment}
In \Cref{sec:uniform} we showed a negative result that for the general class of LSAs, it is not possible to pick a fixed uniform step-size that can be used for any instance of the class. Further,
recall from our discussion in \Cref{sec:mainresults} that as $\alpha\ra\alpha_P$ (or $\alpha\ra\alpha_{P_U}$), the MSE blows up and for $\alpha>\alpha_P$ the LSA is unstable because the spectral norm of the random matrix $I-\alpha A_t$ is no longer less that unity. Combining these facts, we conclude that if the step-size is incorrect for a particular instance then we would observe instability, a fact that can be used to reset the step-size to a smaller value, and this procedure can be continued in an iterative fashion to find the right choice of step-size for a given problem. We formalize this idea in the form of a step-size tuning algorithm in \Cref{alg:tuning}
\end{comment}
\begin{algorithm}[H]
\caption*{Automatic Tuning of Constant Step-Size}
\begin{algorithmic}[1]
\STATE{Initialize: $\theta_0$, $\alpha=1$, $i=1$, $\rho_0=0$}
\FOR{$t=1,2,\ldots, $}
\STATE{$\theta_t=\theta_{t-1}+\alpha (b_t-A_t\theta_{t-1})$}
\IF{$t (mod k)==0$}
\STATE{$\rho_{i}=\norm{\theta_t}$}
\IF{IsUnstable($\{\rho_j,j=1,\ldots,i\}$)==1}
\STATE{$\alpha=\alpha/2$}
\ENDIF
\STATE{$i=i+1$}
\ENDIF
\ENDFOR
\end{algorithmic}
\label{alg:tuning}
\end{algorithm}
\begin{comment}
In \Cref{alg:tuning} IsUnstable$(\rho_j,j=1,\ldots,i\}$) is a routine whose input is $\norm{\theta_t}$ measured at time instances $k$ apart. It is easy to show from Gronwall type arguments that the $\norm{\theta_t}\leq (1-exp{\beta t})$, where $\beta>0$ implies stability and $\beta<0$ imples instability. From this Gronwall based inequality, we observe that when the iterates are stable the ratio $\frac{\rho_{k+1}}{\rho_k},k=1,\ldots,i-1$ should be `roughly' decreasing (because there is the iterates are persistently corrupted by noise) and should converge to $1$. This condition on the sequence $\frac{\rho_{k+1}}{\rho_k}$ can be checked in practice with appropriate tolerances.
\end{comment}
\begin{comment}
In order to calculate the rates we make use of the following asymptotic result by \cite{borkarbook}.
\begin{theorem}\label{th:borkar}
Let $\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1})$, where $\theta_t,b_t\in\R^d$ and $A_t\in\R^{\dcd}$ that satisfy the conditions in \Cref{assmp:lsa} and $\sum_{t\geq 0}\alpha_t=\infty$, $\sum_{t\geq 0}\alpha^2_t<\infty$. Let $t_{ode}(k)=\eqdef \sum_{i=0}^{k-1}\alpha_n$ and let $\tilde{\theta}(t)$ be the trajectory of the ordinary differential equation $\dot{\theta}(t)=\EE{b_t}-A_P\theta(t)$. Then it follows that
\begin{align*}
\lim_{k\ra\infty}\sup\norm{\tilde{\theta_t(t_{ode}(k))}-\theta_t} \ra 0
\end{align*}
\end{theorem}

### 2017-05-13 uniform.tex
refreshing

\section{Uniform stepsizes}
If $\P$ is weakly admissible, then one can choose a stepsize $\alpha_{\P}>0$ solely based on the knowledge of $\P$ and
conclude that no matter the joint distribution $P$ of $(b_t,A_t)$, as long as the distribution of $P$ is an element of $\P$, the conclusions of \Cref{th:rate} hold. Further, if $\P$ is admissible then the error bound stated in  \cref{maintheorem} becomes independent of the instance. Hence, an interesting question to investigate is whether a given set $\P$ is (weakly) admissible. We consider some reasonable classes of distributions and test whether or not they are (weakly) admissible.
\begin{definition}
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item For $B>0$, let $\P_B$ be the set of distributions over $\C_B^{d}\times \C_B^{\dcd}$ where for every $P\in \P_B$ is positive definite.
\item  $\P_{\text{PSD},B}$ be such that  $\P^M_{\text{PSD},B}=\{ P^M \in \P^M_B \,:\,  \supp(P^M)\subset \text{PSD} \}\,$ , where PSD denotes the set of positive semi-definite matrices.
\end{itemize}
\end{definition}
Note that all distributions are a subset of $\P_B$ and in most applications the knowledge of a suitable $B>0$ is available,and by appropriately normalizing the vectors and matrices involved, we may as well assume that $B=1$ without any loss of generality.
\begin{proposition}\label{notwad}
The set $\P_B$ is not weakly admissible.
\end{proposition}
\begin{proposition}
For any $B>0$, the set $\P_{\text{PSD},B}$ is weakly admissible
and in particular any $0<\alpha < 2/B$ witnesses this.
\end{proposition}
\begin{proposition}
For any $B>0$, the set $\P_{\text{PSD},B}$ is not admissible.
\end{proposition}

@@

\begin{comment}
\begin{proof}
Take any $P\in \P_{\text{PSD}}$ and let $H\sim P$.
Consider the SVD of $H$: $H = U \Lambda U^\top$ where $U$ is orthonormal and $\Lambda$ is diagonal with
nonnegative elements. Note that $\Lambda \preceq B\, \I$ and thus $\Lambda^2 \preceq B \Lambda$.
Then for any $x\in \R^d$, $x^\top H^\top H x = x^\top U \Lambda^2 U^\top x \le B x^\top U \Lambda U^\top x = B x^\top H x$.
Taking expectations we find that $x^\top Q_P x \le B x^\top H_P x$.
Hence, $\rho_\alpha(P) = 2 x^\top H_P x - \alpha x^\top Q_P x \ge (2- \alpha B ) \,x^\top H_P x $.
Thus, for any $\alpha<2/B$, $\rho_\alpha(P)>0$.
\end{proof}
\end{comment}
\todoc[inline]{So this altogether is weaker than what Bach founds. They prove strong admissibility.
I am guessing that for this they use that the loss is special and also that the noise is special.
This would be a good place to somehow incorporate this with a (short?) proof.
}
Thus \Cref{notwad} positive definiteness of $P\in \P_B$ alone is not sufficient for weak admissibility. When we impost
In linear least-squares problems (cf. \cref{ex:leastsquares}) the set $\P$ has more structure.
In particular, the random matrices $\{A_t\}$ are PSD. \todoc{The standard notation is $\mathrm{S}^{++}$ for the set of positive definite matrices. Introduce these and use them below..}
We have the following proposition:

>>>>

\section{Uniform stepsizes}
If $\P$ is weakly admissible, then one can choose a stepsize $\alpha_{\P}>0$ solely based on the knowledge of $\P$ and
conclude that no matter the joint distribution $P$ of $(b_t,A_t)$, as long as the distribution of $P$ is an element of $\P$, the conclusions of \Cref{th:rate} hold. Further, if $\P$ is admissible then the error bound stated in  \cref{maintheorem} becomes independent of the instance. Hence, an interesting question to investigate is whether a given set $\P$ is (weakly) admissible. We consider some reasonable classes of distributions and test whether or not they are (weakly) admissible.
\begin{definition}\label{def:pclass}
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item For $B>0$, let $\P_B$ be the set of distributions over $\C_B^{d}\times \C_B^{\dcd}$ where for every $P\in \P_B$ is positive definite.
\item  $\P_{\text{PSD},B}$ be such that  $\P^M_{\text{PSD},B}=\{ P^M \in \P^M_B \,:\,  \supp(P^M)\subset \text{PSD} \}\,$ , where PSD denotes the set of positive semi-definite matrices.
\item  $\P_{\text{PD},B}$ be such that  $\P^M_{\text{SD},B}=\{ P^M \in \P^M_B \,:\,  \supp(P^M)\subset \text{PD} \}\,$ , where PD denotes the set of positive definite matrices.
\end{itemize}
\end{definition}
Note that all distributions are a subset of $\P_B$ and in most applications the knowledge of a suitable $B>0$ is available,and by appropriately normalizing the vectors and matrices involved, we may as well assume that $B=1$ without any loss of generality.
\begin{proposition}\label{notwad}
The set $\P_B$ is not weakly admissible.
\end{proposition}
\begin{proposition}
For any $B>0$, the set $\P_{\text{PSD},B}$ is weakly admissible
and in particular any $0<\alpha < 2/B$ witnesses this.
\end{proposition}
\begin{proposition}
For any $B>0$, the set $\P_{\text{PSD},B}$ is not admissible.
\end{proposition}

@@

\begin{comment}
\begin{proof}
Take any $P\in \P_{\text{PSD}}$ and let $H\sim P$.
Consider the SVD of $H$: $H = U \Lambda U^\top$ where $U$ is orthonormal and $\Lambda$ is diagonal with
nonnegative elements. Note that $\Lambda \preceq B\, \I$ and thus $\Lambda^2 \preceq B \Lambda$.
Then for any $x\in \R^d$, $x^\top H^\top H x = x^\top U \Lambda^2 U^\top x \le B x^\top U \Lambda U^\top x = B x^\top H x$.
Taking expectations we find that $x^\top Q_P x \le B x^\top H_P x$.
Hence, $\rho_\alpha(P) = 2 x^\top H_P x - \alpha x^\top Q_P x \ge (2- \alpha B ) \,x^\top H_P x $.
Thus, for any $\alpha<2/B$, $\rho_\alpha(P)>0$.
\end{proof}
\end{comment}
\todoc[inline]{So this altogether is weaker than what Bach founds. They prove strong admissibility.
I am guessing that for this they use that the loss is special and also that the noise is special.
This would be a good place to somehow incorporate this with a (short?) proof.
}
Thus \Cref{notwad} positive definiteness of $P\in \P_B$ alone is not sufficient for weak admissibility. However, when we impose additional structure in the data in the form of positive semidefiniteness,  we then observe weak admissibility in \Cref{wad}, and at the same time, it is not enough to guarantee admissiblity in \Cref{notad}. Finally, when we impose strucutre in the form of positive definiteness we observe admissibility.
\FloatBarrier
\begin{table}[H]
\begin{tabular}{|c|c|c|c|}\hline
Problem Class & Weak Admissible & Admissible &Example \\ \hline
$\P_B$ &\ding{53} &\ding{53} & RL\\\hline
$\P_{\text{PSD},B}$ &\ding{51} &\ding{53} & SGD for Linear Regression\\\hline
$\P_{\text{PD},B}$ &\ding{51} &\ding{51} & SGD for Linear Regression with regularization\\\hline
\end{tabular}
\end{table}
